---
title: "3 simple ways to get your loan approved"
author: "Richard Kublik"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  html_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      comment = NA,
                      fig.align = "center")

library(readr)
library(dplyr)
library(caret)
library(rpart)
library(rattle)
library(rpart.plot)
library(RColorBrewer)
library(ranger)
library(e1071)
library(pROC)
library(randomForest)
library(glmnet)
library(ggmosaic)
```

# Introduction
In this project we will demonstrate a number of basic machine learning algorithms, and apply them to the [Loan Prediction practice competition](https://datahack.analyticsvidhya.com/contest/practice-problem-loan-prediction-iii/) offered by [Analytics Vidhya](https://www.analyticsvidhya.com/).

## The Problem
Dream Housing Finance company provides all types of home loans. They have a presence across all urban, semi-urban, and rural areas. Prior to the actual loan application, the company validates the customer's eligibility for a loan.

Dream Housing Finance wants to automate the loan eligibility process, and provide a real time decision based on customer details provided through an online application form.

## The Data
Dream Housing Finance collects a number of customer details, These are provided in a .csv file with the columns:

**Variable**      **Description**
-------------     -----------------
Loan_ID           Unique Loan ID
Gender            Male/ Female
Married           Applicant married (Y/N)
Dependents        Number of dependents
Education         Applicant Education (Graduate/ Under Graduate)
Self_Employed     Self employed (Y/N)
ApplicantIncome   Applicant income
CoapplicantIncome Coapplicant income
LoanAmount        Loan amount in thousands
Loan_Amount_Term  Term of loan in months
Credit_History    credit history meets guidelines
Property_Area     Urban/ Semi Urban/ Rural
Loan_Status       Loan approved (Y/N)

# Pre-processing
As with most real world data sets, the loan data provided by Dream Housing Finance requires some initial cleanup prior to working with it. In particular, there are missing values that must be dealt with before any functional model can be created. We begin by loading the training data provided, and converting the character data types to factors.

```{r echo = TRUE, message = FALSE, results = "asis"}
loan_data <- read_csv(file.path("..","data","train.csv")) %>% 
  mutate(Gender = factor(Gender),
         Married = factor(Married),
         Dependents = as.integer(gsub("+", "", Dependents, fixed = TRUE)),
         Education = factor(Education),
         Self_Employed = factor(Self_Employed),
         Credit_History = factor(Credit_History),
         Property_Area = factor(Property_Area),
         Loan_Status = factor(Loan_Status),
         Loan_Amount_Term = factor(Loan_Amount_Term))

```

If we look at the summary statistics of the dataset, we find a number of variables have missing values:
```{r echo = TRUE}
summary(loan_data)
```

For the categorical variables, we will impute the missing values with the most common value:

* **Gender:**  5/6ths of the applicants are male: impute NA's with Male
* **Married:** 4/6ths are married: impute NA's with Yes
* **Self_Employed:** 5/6ths are not self employed: impute NA's with No
* **Loan_Amount_Term:** impute with median (360)
* **Credit_History:** 5/6ths have good credit: impute NA's with 1

We will use a simple function to impute these values.

```{r echo = TRUE}
impute_value <- function(data, column, value){
  data[[column]][which(is.na(data[[column]]))] <- value
  return(data)
}

loan_data <- loan_data %>% 
  impute_value("Gender", "Male") %>% 
  impute_value("Married", "Yes") %>% 
  impute_value("Self_Employed", "No") %>% 
  impute_value("Loan_Amount_Term", 360) %>% 
  impute_value("Credit_History", 1)

```

For the number of `Dependents`, we will group the loan applicants by gender and marital status, and impute the NA's with the median number of dependents within each group.
```{r}
loan_data <- loan_data %>% 
  group_by(Gender, Married) %>% 
  mutate(Dependents = ifelse(is.na(Dependents), 
                             as.integer(median(Dependents, na.rm = TRUE)), 
                             Dependents)) %>% 
  ungroup()
summary(loan_data)
```

For the `LoanAmount` we will group the applications based on `Property_Area` and `Loan_Amount_Term` since it is likely that loan values will depend on the area and loan term.
```{r}
loan_data <- loan_data %>% 
  group_by(Property_Area, Loan_Amount_Term) %>% 
  mutate(LoanAmount = ifelse(is.na(LoanAmount), 
                             as.integer(mean(LoanAmount, na.rm = TRUE)), 
                             LoanAmount)) %>% 
  ungroup()
```

# Data Exploration
Before building a predictive model, we want to explore the data and see if there are any variables that have an obvious correlation to the approval or rejection of a loan application. To do this we will make a series of plots to visualize the relationship between each variable:
```{r echo = FALSE, results = "asis"}

mosaic_plot <- function(data, varx, vary){
  
  stats <- data %>%
    count_(varx)
    
  total_rows <- dim(data)[1]
  data$weight = 0

  for (s in 1:dim(stats)[1]) {
    data$weight[which(data[[varx]] == stats[[1]][[s]])] = stats$n[[s]]/total_rows
  }
  
  data %>% 
    ggplot() +
    geom_mosaic(aes(weight = weight, x = product(get(vary), get(varx)), fill = get(vary))) +
    labs(x = varx,
         y = "Percent",
         title = sprintf("%s vs %s", vary, varx)) + 
    scale_fill_discrete(name = vary)
  
}
```
<div class="row">
<div class="col-md-4 col-sm-6">
```{r echo = FALSE}
mosaic_plot(loan_data, "Credit_History", "Loan_Status")
```
</div><div class="col-md-4 col-sm-6">
```{r echo = FALSE}
mosaic_plot(loan_data,"Gender", "Loan_Status")
```
</div><div class="col-md-4 col-sm-6">
```{r echo = FALSE}
mosaic_plot(loan_data,"Dependents", "Loan_Status")
```
</div><div class="col-md-4 col-sm-6">
```{r echo = FALSE}
mosaic_plot(loan_data,"Married", "Loan_Status")
```
</div><div class="col-md-4 col-sm-6">
```{r echo = FALSE}
mosaic_plot(loan_data,"Education", "Loan_Status")
```
</div><div class="col-md-4 col-sm-6">
```{r echo = FALSE}
mosaic_plot(loan_data,"Self_Employed", "Loan_Status")
```
</div><div class="col-md-4 col-sm-6">
```{r echo = FALSE}
mosaic_plot(loan_data,"Property_Area", "Loan_Status")
```
</div>
</div>


