---
title: "3 simple ways to get your loan approved"
author: "Richard Kublik"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  html_document:
    toc: true
    toc_depth: 2
    toc_float: 
      collapsed: true
      smooth_scroll: true
    
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      comment = NA,
                      fig.align = "center")
#set.seed(42)
```

# Introduction
In this project we will demonstrate a number of basic machine learning algorithms, and apply them to the [Loan Prediction practice competition](https://datahack.analyticsvidhya.com/contest/practice-problem-loan-prediction-iii/) offered by [Analytics Vidhya](https://www.analyticsvidhya.com/).

## The Problem
Dream Housing Finance company provides all types of home loans. They have a presence across all urban, semi-urban, and rural areas. Prior to the actual loan application, the company validates the customer's eligibility for a loan.

Dream Housing Finance wants to automate the loan eligibility process, and provide a real time decision based on customer details provided through an online application form.

## The Data
Dream Housing Finance collects a number of customer details, These are provided in a .csv file with the columns:

**Variable**      **Description**
-------------     -----------------
Loan_ID           Unique Loan ID
Gender            Male/ Female
Married           Applicant married (Y/N)
Dependents        Number of dependents
Education         Applicant Education (Graduate/ Under Graduate)
Self_Employed     Self employed (Y/N)
ApplicantIncome   Applicant income
CoapplicantIncome Coapplicant income
LoanAmount        Loan amount in thousands
Loan_Amount_Term  Term of loan in months
Credit_History    credit history meets guidelines
Property_Area     Urban/Semi Urban/Rural
Loan_Status       Loan approved (Y/N)

# Pre-processing
As with most real world data sets, the loan data provided by Dream Housing Finance requires some initial cleanup prior to working with it. In particular, there are missing values that must be dealt with before any functional model can be created. We begin by loading the training data provided, and converting the character data types to factors.

```{r echo = TRUE, message = FALSE, results = "asis"}
library(dplyr)
library(readr)
library(caret)
library(rpart)
library(rattle)
library(rpart.plot)
library(RColorBrewer)
library(ranger)
library(e1071)
library(pROC)
library(randomForest)
library(glmnet)
library(ggmosaic)
library(ggbiplot)

pre_process <- function(data){
  data %>% 
    mutate(Gender = factor(Gender),
           Married = factor(Married),
           Dependents = as.integer(gsub("+", "", Dependents, fixed = TRUE)),
           Education = factor(Education),
           Self_Employed = factor(Self_Employed),
           Credit_History = factor(Credit_History),
           Property_Area = factor(Property_Area),
           Loan_Amount_Term = factor(Loan_Amount_Term))
}
loan_data <- read_csv(file.path("..","data","train.csv")) %>% 
  pre_process() %>% 
  mutate(Loan_Status = factor(Loan_Status))

```

If we look at the summary statistics of the dataset, we find a number of variables have missing values:
```{r echo = TRUE}
summary(loan_data)
```

For the categorical variables, we will impute the missing values with the most common value:

* **Gender:**  5/6ths of the applicants are male: impute NA's with Male
* **Married:** 4/6ths are married: impute NA's with Yes
* **Self_Employed:** 5/6ths are not self employed: impute NA's with No
* **Loan_Amount_Term:** impute with median (360)
* **Credit_History:** 5/6ths have good credit: impute NA's with 1

For the continuous variables, we will use some groupings to obtain more refined estimates of the missing values:

* **Dependents:** group by gender and marital status, and impute the NA's with the median number of dependents within each group.
* **LoanAmount:** group by `Property_Area` and `Loan_Amount_Term` since it is likely that loan values will depend on the area and loan term.

We will use a simple function to impute these values.

```{r echo = TRUE}
impute_value <- function(data, column, value){
  data[[column]][which(is.na(data[[column]]))] <- value
  return(data)
}

clean <- function(data){
  data %>% 
    # Impute Categorical variables with the most common value
    impute_value("Gender", "Male") %>% 
    impute_value("Married", "Yes") %>% 
    impute_value("Self_Employed", "No") %>% 
    impute_value("Loan_Amount_Term", 360) %>% 
    impute_value("Credit_History", 1) %>% 
    # Impute median number of dependents, by gender and marital status
    group_by(Gender, Married) %>% 
    mutate(Dependents = ifelse(is.na(Dependents), 
                               as.integer(median(Dependents, na.rm = TRUE)), 
                               Dependents)) %>% 
    ungroup() %>% 
    # Impute mean loan amount, by property area and loan term length
    group_by(Property_Area, Loan_Amount_Term) %>% 
    mutate(LoanAmount = ifelse(is.na(LoanAmount), 
                               as.integer(mean(LoanAmount, na.rm = TRUE)), 
                               LoanAmount)) %>% 
    ungroup()
}
loan_data <- clean(loan_data)
```

# Data Exploration
Before building a predictive model, we want to explore the data and see what insight we can gain.

## Visual Inspection
As there are a relatively small number of variables, we can look at each one individually to see if there is an obvious correlation to the approval or rejection of a loan application. To do this we will make a series of plots to visualize the relationship between each variable:
```{r echo = FALSE, results = "asis"}

mosaic_plot <- function(data, varx, vary){
  stats <- data %>%
    count_(varx)
    
  total_rows <- dim(data)[1]
  data$weight = 0

  for (s in 1:dim(stats)[1]) {
    data$weight[which(data[[varx]] == stats[[1]][[s]])] = stats$n[[s]]/total_rows
  }
  
  data %>% 
    ggplot() +
    geom_mosaic(aes(weight = weight, x = product(get(vary), get(varx)), fill = get(vary))) +
    labs(x = varx,
         y = "Percent",
         title = sprintf("%s vs %s", vary, varx)) + 
    scale_fill_discrete(name = vary)
  
}

box_plot <- function(data, varx, vary){
  data %>% 
    ggplot(aes_string(x = varx, y = vary)) + 
    geom_boxplot(varwidth = TRUE) +
    labs(x = varx,
         y = vary,
         title = sprintf("%s vs %s", vary, varx))  
    

}
```
<div class="row">
<div class="col-md-4 col-sm-6">
```{r echo = FALSE}
mosaic_plot(loan_data, "Credit_History", "Loan_Status")
```
</div><div class="col-md-4 col-sm-6">
```{r echo = FALSE}
mosaic_plot(loan_data,"Gender", "Loan_Status")
```
</div><div class="col-md-4 col-sm-6">
```{r echo = FALSE}
mosaic_plot(loan_data,"Dependents", "Loan_Status")
```
</div><div class="col-md-4 col-sm-6">
```{r echo = FALSE}
mosaic_plot(loan_data,"Married", "Loan_Status")
```
</div><div class="col-md-4 col-sm-6">
```{r echo = FALSE}
mosaic_plot(loan_data,"Education", "Loan_Status")
```
</div><div class="col-md-4 col-sm-6">
```{r echo = FALSE}
mosaic_plot(loan_data,"Self_Employed", "Loan_Status")
```
</div><div class="col-md-4 col-sm-6">
```{r echo = FALSE}
mosaic_plot(loan_data,"Property_Area", "Loan_Status")
```
</div><div class="col-md-4 col-sm-6">
```{r echo = FALSE}
mosaic_plot(loan_data,"Loan_Amount_Term", "Loan_Status")
```
</div><div class="col-md-4 col-sm-6">
```{r echo = FALSE}
box_plot(loan_data, "Loan_Status", "LoanAmount")
```
</div><div class="col-md-4 col-sm-6">
```{r echo = FALSE}
box_plot(loan_data, "Loan_Status", "ApplicantIncome")
```
</div><div class="col-md-4 col-sm-6">
```{r echo = FALSE}
box_plot(loan_data, "Loan_Status", "CoapplicantIncome")
```
</div><div class="col-md-4 col-sm-6">
```{r echo = FALSE}
loan_data %>% 
  mutate(TotalIncome = ApplicantIncome + CoapplicantIncome) %>% 
  box_plot("Loan_Status", "TotalIncome")
```
</div>
</div>

In the final plot, `TotalIncome` is a new variable computed as the sum of `ApplicantIncome` and `CoapplicantIncome`.

From the plots we can draw a number of conclusions:

* Credit history shows the most significant correlation with loan approval. We will include this in all of our models
* Marital status, Education level, and Property area appear to have a small correlation with loan approval
* The remaining variables do not appear to have any significant correlation with loan approval.

## Principal Component Ananlysis
In many cases, Principal Component Analysis (PCA) provides some insight into the data. By plotting the data in PCA space we can potentially see clustering in the data, along with information about co-linearity of the variables. Plotting the PCA of the input variables, we obtain:
```{r echo = TRUE}
train_pca <- loan_data %>% 
  select(-Loan_ID, -Loan_Status) %>%
  # convert factors to numeric values
  mutate(Gender = as.numeric(Gender),
         Married = as.numeric(Married),
         Education = as.numeric(Education),
         Self_Employed = as.numeric(Self_Employed),
         Loan_Amount_Term = as.numeric(Loan_Amount_Term),
         Credit_History = as.numeric(Credit_History),
         Property_Area = as.numeric(Property_Area)) %>% 
  # Calculate PCA
  prcomp(center = TRUE, scale = TRUE) 
  
ggbiplot(train_pca) 

```

This plot shows all of the data points in PCA space, and includes arrows mapping the variables to the 2 dimensional PCA space. Though the text is a bit difficult to read, we see that:

* `Married` and `Gender` are colinear
* `CoapplicantIncome` and `Dependents` are colinear
* `ApplicantIncome` and `Self_Employed` are colinear

Using this information we will be able to simplify the models that we build. Additionally, when we are building the logistic model, we will want to avoid having variables that are co-linear.

```{r include = FALSE}
# remove plyr package, as the conflicts with dplyr cause problems later
detach("package:ggbiplot", unload = TRUE)
detach("package:plyr")
```

# Predictive Models
Using the insight we have gained in the previous section, we will construct a number of predictive models and compare their performance. To insure a fair comparison, we will split the provided data into test and training sets and use these to evaluate how well our models do on unseen data. Using the `caret` package, we create `loan$train`, and `loan$test`:
```{r echo = TRUE}
train_idx = createDataPartition(loan_data$Loan_Status, p = 0.7, list = FALSE)
loan = list(train = loan_data[train_idx,],
            test = loan_data[-train_idx,])
```

In addition, since the data was provided as part of an online competition, we can submit the results of our models to see how they compare to models developed by other people. We will pre-process the test data set using the functions we created earlier:
```{r echo = TRUE, message = FALSE}
loan_test_data <- read_csv(file.path("..","data","test.csv")) %>% 
  pre_process() %>% 
  clean()
```

## Credit History as Sole Predictor
From our earlier exploration, we saw that the `Credit_History` variable has the most obvious correlation with `Loan_Status`. For this initial attempt at a model, we will predict loan approval for all customers with `Credit_History = 1`. Applying this to our test partition
```{r echo = TRUE}
credit_history_success <- loan$test %>% 
  mutate(loan_approved = Loan_Status == "Y",
         credit_only = Credit_History == 1) %>% 
  summarize(n = n(),
            success = sum(credit_only)/n)

```
we obtain an accuracy of `r round(100*credit_history_success$success[1])`%, which seems quite good. However, this is likely dependent on the way our training and testing data sets were partitioned. 

If we apply this same technique to the provided test data and submit it to the competition (using the code provided below), we obtain an accuracy of 78% which puts us in 945th place (tied with 319 other people at the time of writing).

```{r echo = TRUE}
loan_test_data %>% 
  mutate(Loan_Status = ifelse(Credit_History == 1, "Y", "N")) %>% 
  select(Loan_ID, Loan_Status) %>%  
  write_csv(file.path("credit_only.csv"))
```

## Logistic Regression

For our logistic regression solution, we will include a few more variables. From our visual inspection of the variables, and the results of PCA, we will build a linear regression model that includes the variables: `Credit_History`, `Married`, `Education`, `Property_Area`. Using the `caret` package to do cross-validation, we train our model as shown below:

```{r echo = TRUE, message = FALSE}

# Define a common training control:
train_control <- trainControl(method = "repeatedcv",
                              number = 10,
                              repeats = 10,
                              summaryFunction = twoClassSummary,
                              classProbs = TRUE,
                              verboseIter = FALSE)

logit_model <- train(Loan_Status ~ Credit_History + Married + Education + Property_Area,
                   data = loan$train,
                   method = "glm",
                   metric = "ROC",
                   trControl = train_control,
                   family = binomial(link = "logit"))
```

If we test our model on our training partition

```{r echo = TRUE}
logit_pred <- predict(logit_model, loan$test, type = "prob")
logit_accuracy <- data.frame(loan_id = loan$test$Loan_ID,
                       loan_granted = loan$test$Loan_Status == "Y",
                       logit = logit_pred$Y > 0.5) %>%
  summarize(n = n(),
            accuracy = sum(logit == loan_granted)/n)

round(logit_accuracy$accuracy[1]*100)
```

we obtain an accuracy of `r round(logit_accuracy$accuracy[1]*100)`%. We can also examine the variables included in the model to determine which are the most significant.
```{r echo = TRUE}
summary(logit_model)$coefficients
```
Here we see that the variables with the most significance (smallest p-value) are `Credit_History` and `Property_Area`. We know that including too many variables in a model can lead to over-fitting, and poor performance on new data. Let's look at a model that only contains the two most important variables.  Following the same steps as above, we will train and test a reduced logistic model.

```{r echo = TRUE, message=FALSE, warning=FALSE}
logit_reduced_model <- train(Loan_Status ~ Credit_History  + Property_Area,
                   data = loan$train,
                   method = "glm",
                   metric = "ROC",
                   trControl = train_control,
                   family = binomial(link = "logit"))
logit_reduced_pred <- predict(logit_reduced_model, loan$test, type = "prob")
logit_reduced_accuracy <- data.frame(loan_id = loan$test$Loan_ID,
                       loan_granted = loan$test$Loan_Status == "Y",
                       logit = logit_reduced_pred$Y > 0.5) %>%
  summarize(n = n(),
            accuracy = sum(logit == loan_granted)/n)
summary(logit_reduced_model)$coefficients
round(logit_reduced_accuracy$accuracy[1]*100)
```

This model gives us an accuracy of `r round(logit_reduced_accuracy$accuracy[1]*100)`% on our test partition. Our reduced model performs just as well as the more detailed model, however, both models are out-performed on our training partition by the credit history only model. The final test will be how well our logistic model performs on the provided test data.

We will use the reduced model for our submission. We will train the model using the full training data set, and then generate predictions to submit.

```{r echo = TRUE}
logit_submission_model <- train(Loan_Status ~ Credit_History  + Property_Area,
                   data = loan_data,
                   method = "glm",
                   metric = "ROC",
                   trControl = train_control,
                   family = binomial(link = "logit"))

logit_submission_pred <- predict(logit_reduced_model, loan_test_data, type = "prob")
data.frame(Loan_ID = loan_test_data$Loan_ID,
           Loan_Status = ifelse(logit_submission_pred$Y > 0.5, "Y", "N")) %>% 
    write_csv(file.path("logistic_regression.csv"))
```

When we submit this result, our reported accuracy is again 78%. It seems that Logistic regression isn't the right approach for this problem.

## Decision Tree

We will again take the varibles determined to be the most important and use them in our decision tree model. Again, we will utilize the `caret` package to perform the cross validation on our model.
```{r echo = TRUE}

rpart_model <- train(#Loan_Status ~ Credit_History + Married + Dependents + Education + ApplicantIncome + CoapplicantIncome + LoanAmount + Property_Area,
                     Loan_Status ~ Credit_History + Married + Education + Property_Area,
                     data = loan$train,
                     method = "rpart",
                     metric = "ROC",
                     trControl = train_control)

```




```{r echo = FALSE}
fancyRpartPlot(rpart_model$finalModel)


```
